---
title: "Data Modeling Preparation"
format: pdf
---

## Environment Setup

```{r setup}
#| include: false
#| warning: false
#| message: false

library(tidyverse)      
library(arrow)       
```

## Load Data

```{r}
loans_NY <- read_parquet(here("data", "02-analysis_data", "loans_NY_clean.parquet"))

# SELECT MODELING FEATURES
# Isolate only the variables required for the model

loans_model_data <- loans_NY |>
  select(
    denied, race, applicant_poc, sex, age_bracket, 
    log_income, dti_bracket, combined_loan_to_value_ratio, 
    log_loan_amount, tract_minority_population_percent, 
    tract_to_msa_income_percentage, lender_id, county_code, census_tract)

```


## Preprocessing

```{r}
# SET FACTOR BASELINES & STANDARDIZE CONTINUOUS VARIABLES

# Scaling continuous predictors for appropriate convergence and comparison

model_data <- loans_model_data |>
  mutate(
    # Remove ordered status for R models
    age = factor(age_bracket, ordered = FALSE),
    dti = factor(dti_bracket, ordered = FALSE),

    # Standardize all continuous predictors
    log_income_z = scale(log_income)[, 1],
    log_loan_amount_z = scale(log_loan_amount)[, 1],
    ltv_z = scale(combined_loan_to_value_ratio)[, 1],
    minority_pct_z = scale(tract_minority_population_percent)[, 1],
    msa_income_pct_z = scale(tract_to_msa_income_percentage)[, 1]
  )

# EXPORT MODEL-READY DATA
saveRDS(model_data, here("data", "02-analysis_data", "model_data.rds"))
```


## Correlation Analysis

```{r}
# Correlation Matrix
continuous_vars <- model_data |>
  select(log_income_z, log_loan_amount_z, ltv_z, minority_pct_z, msa_income_pct_z)
  
cor_matrix <- cor(continuous_vars, use = "complete.obs")

round(cor_matrix, 2)
```

Prior to model fitting, we assessed the continuous predictors for multicollinearity. A high Pearson correlation (r = 0.77) was observed between the log-transformed income and loan amount variables. Because including highly correlated predictors will weaken coefficient interpretation, loan amount was dropped as a covariate predicor for a better model.

